\documentclass{uofsthesis-cs}
\input{preamble}

\title{Braid groups and Baxter polynomials}
\author{Noah Friesen}
\degree{\MSc}
\defencedate{July 2024}
\department{Mathematics and Statistics}

\ptuaddress{Head of the Department of Mathematics and Statistics\\
University of Saskatchewan\\
142 McLean Hall, 106 Wiggins Road\\
Saskatoon, Saskatchewan S7N 5E6 Canada
}

\abstract{
It is well known that the braid group of a simple Lie algebra acts on its integrable representations via products of exponentials of its Chevalley generators.
In particular, the Yangian is an integrable representation, so there is an action of the braid group on this space.
We show that modifying this action induces an action of the braid group on a certain commutative subalgebra of the Yangian by Hopf algebra automorphisms.
By dualizing this modified action, we recover an action of the braid group on tuples of rational functions defined in the work of Y. Tan.
Using this dual action, we prove a conjecture of S. Gautam and C. Wendlandt that the two sufficient conditions for the tensor product of finite-dimensional irreducible representations of the Yangian to be cyclic are identical.
One of these conditions involves the aforementioned action of the braid group on rational functions, and the other involves roots of the Baxter polynomials, which have many interesting properties and ties to mathematical physics.
}

\declaration{
The content of this thesis is, to the best of my knowledge, original except where attributed to others within the bibliography.

Portions of Chapters \ref{chap:braidgroup}, \ref{chap:weights}, and \ref{chap:baxter} have previously appeared in joint work with my supervisors Alex Weekes and Curtis Wendlandt under the title ``Braid group actions, Baxter polynomials, and affine quantum groups'' (arXiv:2401.06402, to appear in Transactions of the American Mathematical Society; \cite{friesen_braid_2024}), to which I contributed significantly.
}

\acknowledgements{
I would like to thank my supervisors, Prof. Alex Weekes and Prof. Curtis Wendlandt, for their invaluable guidance over the past two years.
I am extremely grateful for their support and this thesis would not have been possible without their direction and feedback.
I would also like to thank Prof. Steven Rayan for his exceptional instruction during my coursework.

I gratefully acknowledge the financial support I have received from the Natural Sciences and Engineering Research Council of Canada (NSERC) through the Canada Graduate Scholarships (CGS M) program.

Finally, I thank my family for their love and support.
}


\begin{document}

\maketitle
\frontmatter

\input{chapters/1-introduction}
\input{chapters/2-yangians}
\input{chapters/3-braidgroup}
\input{chapters/4-weights}
\input{chapters/5-baxter}

% \nocite{*}
\uofsbibliography[alpha]{refs}

% \uofsappendix
% \begin{appendices}
%     \input{chapters/appendix}
% \end{appendices}

\end{document}
